---
title: "Statistical Analysis"
description: |
  Using Univariate Conditional Uniform Graph Tests
preview: gtrans.png
categories:
  - network analytics
  - statistics
  - CUG test
  - statnet
author:
  - name: Kristina Becvar
    url: http://gratefulnetwork.live/
    affiliation: UMass DACSS Program (My Academic Blog Link)
    affiliation_url: https://kristinabecvar.com
date: 2022-05-10
output:
  distill::distill_article:
    toc: true
    code_folding: true

---  


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r echo=FALSE, results=FALSE}

library(ggplot2) 
library(readr) 
library(statnet)
library(ggraph)
library(tidyverse)
library(ergm)
library(dplyr)
library(corrr)
library(magrittr)

suppressWarnings(expr)

```

# Loading Networks

I previously analyzed the network using igraph and statnet, and need to make a decision about which package serves the network best. 

## Affiliation Matrix

Loading the dataset and creating the network to begin my analysis:

```{r code_folding=TRUE}
gd_affiliation <- read.csv('gd_affiliation_matrix.csv', row.names = 1, header = TRUE, check.names = FALSE)
gd_matrix <- as.matrix(gd_affiliation)
```

## Bipartite Projection

```{r code_folding=TRUE}
gd_projection <- gd_matrix%*%t(gd_matrix)
write.csv(gd_projection, file = "gd_projection.csv")
```

## Create statnet Object

```{r echo=FALSE}
set.seed(11)
gd_statnet <- as.network(gd_projection,
               directed = FALSE, 
               bipartite = FALSE,
               loops = FALSE,
               connected = TRUE)
```

# Centrality and Node Data

I am going to load the data frame I saved from the statnet package analysis as well as a dataframe with key comparable results from each package

```{r code_folding=TRUE}
gd_stat_nodes <- read.csv("gd_stat_nodes.csv")
gd_compare <- read.csv("gd_comparison_packages.csv")
```

# Comparing Network Properties

In the semester assignment, we looked at what to do if we are interested in testing whether or not a specific network property, such as transitivity, is higher in the network we are interested in than we would expect from a random network. To test our network against a null hypothesis, we can use the function cug.test() and set the parameters on which we plan to condition. Possible parameters for this function include: size, edges, and dyad.census for sna/statnet.

## Create t-stat Function

```{r code_folding=TRUE}
cug.t<-function(cug.object){
  (cug.object$obs.stat-mean(cug.object$rep.stat))/sd(cug.object$rep.stat)
}
```

## Network Transitivity to Null

Comparing transitivity to null conditional on size

```{r code_folding=TRUE}
#compare network transitivity to null conditional on size
trans.cug<-cug.test(gd_projection,FUN=gtrans,mode="graph",cmode="size")
trans.cug
```

### Plot Results

```{r code_folding=TRUE}
plot(trans.cug)
```

### Inspect Details

```{r code_folding=TRUE}
#inspect CUG test
trans.cug
```

### T-Stat Test

```{r code_folding=TRUE}
#inspect CUG test
cug.t(trans.cug)
```

### Evaluation

There is a reasonable probability that the observed network transitivity of 0.524 could be randomly generated, conditional only on the size of the network. In this case, I cannot accept the alternative hypothesis that the observed transitivity is higher than would be expected from a random network. In fact, the probability that X <= 0.524 is 19.6% and the probability that x >= 0.524 is 80.4%. The t-stat evaluation tells us that the observed transitivity value is less than 1 standard error away from what we would expect, on average.

## Network Degree Centralization to Null

Comparing network degree centralization to null conditional on size

```{r code_folding=TRUE}
#compare network degree centralization to null conditional on size
c.degree.cug <-cug.test(gd_projection,FUN=centralization,  FUN.arg=list(FUN=degree, cmode="degree"), mode="graph", cmode="size")
```

### Plot Results

```{r code_foldin=TRUE}
#plot vs simulation results
plot(c.degree.cug)
```

### Inspect Details

```{r code_folding=TRUE}
#inspect CUG test
c.degree.cug
```

### T-Stat Test

```{r code_folding=TRUE}
#t-stat between observed and simulated networks
cug.t(c.degree.cug)
```

### Evaluation

There is an extremely remote probability that the observed network degree centrality of 5.21 could be randomly generated, conditional only on the size of the network. In this case, I can accept the alternative hypothesis that the observed network degree centrality is higher than would be expected from a random network. .

## Betweenness Centrality

Comparing network betweenness centralization to null conditional on size

```{r code_folding=TRUE}
#compare network betweenness centralization to null conditional on size
b.degree.cug <-cug.test(gd_projection,FUN=centralization,  FUN.arg=list(FUN=betweenness, cmode="undirected"), mode="graph", cmode="size", reps=100) 
```

### Plot Results

```{r code_folding=TRUE}
#plot vs simulation results
plot(b.degree.cug)
```

### Inspect Details

```{r code_folding=TRUE}
#inspect CUG test
b.degree.cug
```

### T-Stat Test

```{r code_folding=TRUE}
#t-stat between observed and simulated networks
cug.t(b.degree.cug)
```

### Evaluation

Again, there is an extremely remote probability that the observed betweenness centrality of 0.376 could be randomly generated, conditional only on the size of the network. In this case, I can accept the alternative hypothesis that the observed network degree centrality is higher than would be expected from a random network. This makes sense with the rest of my network evaluations.

## Transitivity

Comparing transitivity to null conditional on dyad

```{r code_folding=TRUE}
t.degree.cug <-cug.test(gd_projection,FUN=gtrans, mode="graph", cmode="dyad", reps=100) 
```

### Plot Results

```{r code_folding=TRUE}
#plot vs simulation results
plot(t.degree.cug)
```

### Inspect Details

```{r code_folding=TRUE}
#inspect CUG test
t.degree.cug
```

### T-Stat Test

```{r code_folding=TRUE}
#t-stat between observed and simulated networks
cug.t(t.degree.cug)
```

### Evaluation

Unlike the first evaluation of transitivity on size, this evaluation of transitivity on dyad gives a result that is more in line with subsequent calculations.

## Multiple Evaluations

For my research in particular, I want to look at the measures of centrality and through the different modes of evaluation.

```{r code_folding=TRUE}
cug_cent_size <- cug.test(gd_statnet,
                       centralization,
                       FUN.arg=list(FUN=degree), 
                       mode="graph", 
                       cmode="size")
cug_cent_edge <- cug.test(gd_statnet,
                        centralization,
                        FUN.arg=list(FUN=degree), 
                        mode="graph", 
                        cmode="edges")
cug_cent_dyad <- cug.test(gd_statnet,
                       centralization,
                       FUN.arg=list(FUN=degree), 
                       mode="graph", 
                       cmode="dyad.census")

cug_cent_dyad
```

### Visualization

```{r code_folding=TRUE}
degree_centralization <- c(cug_cent_size$obs.stat,
                        cug_cent_edge$obs.stat, 
                        cug_cent_dyad$obs.stat)

rt_tail <- c(cug_cent_size$pgteobs, 
             cug_cent_edge$pgteobs, 
             cug_cent_dyad$pgteobs)

lt_tail <- c(cug_cent_size$plteobs, 
             cug_cent_edge$plteobs, 
             cug_cent_dyad$plteobs)

centrality <- cbind(degree_centralization,
                     rt_tail, 
                     lt_tail)

rownames(centrality) <- c("Size", "Edges", "Dyads")

par(mfrow=c(1,3))
plot(cug_cent_size, main="Centrality \nConditioned on Size" )
plot(cug_cent_edge, main="Centrality \nConditioned on Edges" )
plot(cug_cent_dyad, main="Centrality \nConditioned on Dyads" )
```





